---
title: 'Investigate pval in montoro_droplet: nmf applied to log-transformed data'
author: "Zihao Wang"
date: "6/24/2019"
output: html_document
params:
  dataname: "montoro_droplet"
  K: 5
---

## Goal
* Previously I plotted the p value of fit on GTEx and montoro_droplet dataset. The result is strange for GTEx, with almost all of the p values very close to 1. 

* Here I fit NMF on log-transformed data: $Y = log(X + 1)$ with $X$ the original count. Then $Y \approx W H$. The fit is then $\Lambda  = \exp(WH) - 1$. Then I plot the p values for $X \sim Pois(\Lambda)$ 


## set up environment
```{r setup}
dataname = params$dataname
K = params$K

cat(sprintf("setting:\n"))
cat(sprintf("dataset: %s; K : %d\n", dataname,K))
  

library(readr)
library(NNLM)
library(rsvd)
source("../code/betanmf.R")
source("../code/misc.R")

data.dir = "../bigdata"
out.dir = "../bigdata"

factor.out.file = sprintf("%s_log_factors_betanmf_K%d.csv", dataname,K)
loading.out.file = sprintf("%s_log_loadings_betanmf_K%d.csv", dataname,K)

log_transform <- function(X){
  return(log(1+X))
}

exp_transform <- function(logX){
  return(exp(logX) -  1)
}

pvals_cts <- function(X,Lam){
    set.seed(123)
    n = nrow(X)
    p = ncol(X)
    C = matrix(runif(n*p), nrow = n)
    pvals = C * ppois(X-1, Lam) + (1-C) * ppois(X, Lam)
    return(pvals)
}
```

## load and transform data
```{r load and transform data}
dataname = file.path(data.dir,sprintf("%s.csv", dataname))
counts = read.csv.matrix(dataname)
cat(sprintf("loaded counts: n = %d; p = %d\n", nrow(counts), ncol(counts)))
cat(sprintf("Proportion of counts that are non-zero: %0.1f%%.\n",
            100*mean(counts > 0)))

counts_log = log_transform(counts)
```


## fit data with nmf
```{r fit data with nmf}
cat(sprintf("number of topics to fit: %d\n", K))
set.seed(123)
cat(sprintf("initialization with NNLM::nnmf\n"))
init = nnmf(counts_log,K,method = "scd",loss = "mkl",rel.tol = 1e-8,
              n.threads = 0,max.iter = 2,inner.max.iter = 4,trace = 1,
              verbose = 2)
cat(sprintf("fit with betanmf\n"))
start = proc.time()
fit = betanmf(counts_log,init$W, init$H, verbose = T, eval_every = 10, numiter = 1)
runtime = proc.time() - start
cat(sprintf("fit with runtime %f\n", runtime[[3]]))
```

```{r save fitted model}
F = t(fit$B)
L = fit$A
factor.out.file = file.path(out.dir,factor.out.file)
loading.out.file = file.path(out.dir,loading.out.file)
write_csv(as.data.frame(F),factor.out.file,col_names = FALSE)
write_csv(as.data.frame(L),loading.out.file,col_names = FALSE)
```




## transform back and plot p value
```{r transform fit and plot pval}
Lamhat = fit$A %*% fit$B
Lam = exp_transform(Lamhat)
pval = pvals_cts(counts, Lam)
hist(pval,breaks = 100, main = "p value of fit")
```

### see what are those big p values
```{r see what those big pvals are}
threshold = 0.999
pval_bin = pval > threshold
cat(sprintf("percentage of those have p value > %.3f: %f",threshold, sum(pval_bin)/(nrow(pval)*ncol(pval))))
hist(colSums(pval_bin), breaks = 100, xlab = sprintf("# of pval > %f in a gene", threshold), 
     main = sprintf("number of p values > %f in a gene", threshold))

hist(rowSums(pval_bin), breaks = 100, xlab = sprintf("# of pval > %f in a sample", threshold), 
     main = sprintf("number of p values > %f in a sample", threshold))

```


### Find eigenvalue of binary matrix
```{r find eigenvalues of binary matrix}
set.seed(12345)
svd_out = rsvd(pval_bin,k = 50)
plot(svd_out$d, xlab = "rank of pc", ylab = "d", 
     main = sprintf(" eigenvalues of pval_binary matrix for pvalue > %.3f", threshold))

print("see the scale of the eigenvalues")
svd_out$d
```




