---
title: 'Investigate pval in montoro_droplet: nmf applied to log-transformed data'
author: "Zihao Wang"
date: "6/24/2019"
output: html_document
params:
  dataname: "test"
  K: 5
  log  : TRUE
---

## Goal
* Previously I plotted the p value of fit on GTEx and montoro_droplet dataset. The result is strange for GTEx, with almost all of the p values very close to 1. 

* Here I fit NMF on log-transformed data: $Y = log(X + 1)$ with $X$ the original count. Then $Y \approx W H$. The fit is then $\Lambda  = \exp(WH) - 1$. Then I plot the p values for $X \sim Pois(\Lambda)$ 


## set up environment
```{r setup}
dataname = params$dataname
K = params$K
is_log = params$log

cat(sprintf("setting:\n"))
cat(sprintf("dataset: %s; K : %d; fit_with_log?: %d\n", dataname,K,is_log))

library(readr)
library(rsvd)
source("../code/betanmf.R")
source("../code/misc.R")

data.dir = "../bigdata"

factor.out.file = sprintf("%s_log_factors_betanmf_K%d.csv", dataname,K)
loading.out.file = sprintf("%s_log_loadings_betanmf_K%d.csv", dataname,K)

exp_transform <- function(logX){
  return(exp(logX) -  1)
}

pvals_cts <- function(X,Lam){
    set.seed(123)
    n = nrow(X)
    p = ncol(X)
    C = matrix(runif(n*p), nrow = n)
    pvals = C * ppois(X-1, Lam) + (1-C) * ppois(X, Lam)
    return(pvals)
}
```

## load  data
```{r load data}
dataname = file.path(data.dir,sprintf("%s.csv", dataname))
counts = read.csv.matrix(dataname)
cat(sprintf("loaded counts: n = %d; p = %d\n", nrow(counts), ncol(counts)))
cat(sprintf("Proportion of counts that are non-zero: %0.1f%%.\n",
            100*mean(counts > 0)))
```


## load and compute fitted model
```{r load and compute fitted model}
factor.out.file  <- file.path(data.dir,factor.out.file)
loading.out.file <- file.path(data.dir,loading.out.file)
F0                 <- read.csv.matrix(factor.out.file)
L0                 <- read.csv.matrix(loading.out.file)

Lam = L0 %*% t(F0)

if(is_log){Lam = exp_transform(Lam)}
```


## plot p value
```{r transform fit and plot pval}
pval = pvals_cts(counts, Lam)
hist(pval,breaks = 100, main = "p value of fit")
```

### see what are those big p values
```{r see what those big pvals are}
threshold = 0.999
pval_bin = pval > threshold
cat(sprintf("percentage of those have p value > %.3f: %f",threshold, sum(pval_bin)/(nrow(pval)*ncol(pval))))
hist(colSums(pval_bin), breaks = 100, xlab = sprintf("# of pval > %f in a gene", threshold), 
     main = sprintf("number of p values > %f in a gene", threshold))

hist(rowSums(pval_bin), breaks = 100, xlab = sprintf("# of pval > %f in a sample", threshold), 
     main = sprintf("number of p values > %f in a sample", threshold))

```


### Find eigenvalue of binary matrix
```{r find eigenvalues of binary matrix}
set.seed(12345)
svd_out = rsvd(pval_bin,k = 50)
plot(svd_out$d, xlab = "rank of pc", ylab = "d", 
     main = sprintf(" eigenvalues of pval_binary matrix for pvalue > %.3f", threshold))

print("see the scale of the eigenvalues")
svd_out$d
```




